---
title: "Merge_datacollation_2025"
output: html_document
date: "2025-03-03"
---

The objectives of this script are to clean and collate shiny app academic camera data submissions with government secure and SPI data sources in one workflow, producing two master files: One master file for species detections rarified to 30 minute independent sampling, and one master file for camera locations across all projects. 

I will first test this data on original submissions, but will incorporate cleaned files up to October 2024 and use this script to clean and merge any new data submissions.

Later steps may include integration of DNA and incidental sighting records and basic mapping scripts for species detections by year

Link to github account
```{r}
git config --global user.name "sidarlin"
git config --global user.email "siobhan.darlington@gov.bc.ca"

git remote add origin https://github.com/sidarlin/MMP_DataCollation_2025.git

```


1. Load libraries

```{r setup, include=FALSE}
library(terra)
library(ggplot2)
library(sf)
library(readr)
library(farver)
library(ggplot2)
library(rmapshaper)
library(stars)
library(tidyterra)
library(dplyr)
library(rmapshaper)
library(readxl)
library(dplyr)
library(lubridate)

```


Based on Cindy's cleaned data files, the column names for the camera trap effort data should be:
Project_name
Station_ID
Latitude
Longitude
Camera_make
Camera_model
Bait_lure
Camera_comments
Start_Deployment_date
End_Deployment_date
CamProblem_From
CamProblem_to
Habitat_type
Target_feature
Comments

Species column names should be:
Station_ID
Species
Count
Date_Time
Project_name
Counts


```{r}
knitr::opts_knit$set(echo=TRUE, root.dir = "C:/LocalR/Merge_datacollation_2025/Input_project_sheets")
# setwd("C:/LocalR/Merge_datacollation_2025/Input_project_sheets")
getwd()

file.list <- list.files(path = "C:/LocalR/Merge_datacollation_2025/Input_project_sheets", 
                        pattern = "*.xlsx", full.names = TRUE)
master_data <- list()

## Define column mapping (original names → standard names)
column_mapping <- c(
  "Deployment.Location.ID" = "Station_ID",
  "Site" = "Station_ID",
  "site" = "Station_ID",
  "placename" = "Station_ID",
  "location"  = "Station_ID",
  "latitude" = "Latitude",
  "longitude" = "Longitude",
  "Latitude_DD" = "Latitude",
  "Latitude _DD" = "Latitude",
  "Longitude_DD"= "Longitude",
  "UTM_Northing" = "UTM_N",
  "UTM_N" = "UTM_N",
  "UTM_Easting" = "UTM_E",
  "UTM_E" = "UTM_E",
  "UTM_Zone" = "zone",
  "Camera.Deployment.Begin.Date" = "Start_Deployment_date",
  "Deploy_Date"  = "Start_Deployment_date",
  "start_date" = "Start_Deployment_date",
  "Camera.Deployment.End.Date" = "End_Deployment_date",
  "Retrieval_Date" = "End_Deployment_date",
  "end_date" = "End_Deployment_date",
  "Bait.Type" = "Bait_lure",
  "Feature.Type" = "Target_feature",
  "feature_type" ="Target_feature",
  "Number.of.Animals" = "Count",
  "animal_count" = "Count",
  "Count1" = "Count",
  "total" = "Count", 
  "individual_count" = "Count", 
  "Number" = "Count",
  "sp" = "Species",
  "Species1" = "Species",
  "species_common_name" = "Species",
  "common_name" = "Species",
  "species" = "Species",
  "Date_Time.Captured" = "Date_Time",
  "datetime" = "Date_Time",
  "Datetime" = "Date_Time",
  "image_date_time" = "Date_Time", 
  "dt" = "Date_Time",
  "timestamp" = "Date_Time"
)

## Function to standardize column names
standardize_columns <- function(df) {
  colnames(df) <- recode(colnames(df), !!!column_mapping) # Rename using mapping
  return(df)
}

# Loop through each file
for (i in seq_along(file.list)) {
  file <- file.list[i]
  
  df_cam <- read_excel(file, sheet = 1) %>% standardize_columns()
  df_sp <- read_excel(file, sheet = 2) %>% standardize_columns()
  
  # Extract and clean project name
  project_name <- gsub("^.*data_(.*?)\\-.*$", "\\1", basename(file))
  
 
  # Ensure Station ID is a character
  df_cam$Station_ID <- as.character(df_cam$Station_ID)
  df_sp$Station_ID <- as.character(df_sp$Station_ID)
  
  # Ensure Count is numeric
  df_sp$Count <- as.numeric(df_sp$Count)
  

  # Convert UTM to Latitude/Longitude if UTM columns exist
  if (all(c("UTM_E", "UTM_N", "zone") %in% colnames(df_cam))) {
    df_cam <- df_cam %>%
      filter(!is.na(UTM_E) & !is.na(UTM_N) & !is.na(zone)) %>%
      mutate(zone = as.integer(zone)) %>%
      rowwise() %>%
      mutate(
        geometry = list(st_point(c(UTM_E, UTM_N))),
        crs_utm = 32600 + zone
      ) %>%
      ungroup()
    
    # Convert to sf object
    df_cam_sf <- st_as_sf(df_cam, coords = c("UTM_E", "UTM_N"), crs = unique(df_cam$crs_utm))

    # Transform to WGS84
    df_cam_sf <- st_transform(df_cam_sf, crs = 4326)

    # Extract lat/lon
    df_cam$Longitude <- st_coordinates(df_cam_sf)[, 1]
    df_cam$Latitude <- st_coordinates(df_cam_sf)[, 2]
  }
  
   # Ensure Latitude and Longitude are numeric
  df_cam$Latitude <- as.numeric(df_cam$Latitude)
  df_cam$Longitude <- as.numeric(df_cam$Longitude)
  
  # Standardize deployment data
  deployment_data <- df_cam %>%
    select(any_of(c("Station_ID", "Latitude", "Longitude", "Start_Deployment_date", 
                    "End_Deployment_date", "Camera_make", "Camera_model", "Camera_height", "Bait_lure", "Target_feature"))) %>%
    mutate(across(c(Start_Deployment_date, End_Deployment_date), ~ suppressWarnings(as.Date(.))),
           Project_name = project_name)
  
  # Standardize species data
  species_data <- df_sp %>%
    select(any_of(c("Station_ID", "Species", "Count", "Date_Time"))) %>%
    mutate(Project_name = project_name, # Assign cleaned project name
           Species = gsub("\\.", " ", Species))  # Replace periods with spaces in Species column
  
            # Date_Time = as.POSIXct(Date_Time, tz = "UTC", format = "%Y-%m-%d %H:%M:%S"))

  ## fix Date_Time formatting
if ("Date_Time" %in% colnames(species_data)) {
  species_data <- species_data %>%
    mutate(Date_Time = case_when(
      grepl("AM|PM", Date_Time, ignore.case = TRUE) ~ 
        suppressWarnings(parse_date_time(Date_Time, orders = c("mdy HMS p", "mdy HM p", "ymd HMS p", "ymd HM p"))),
      TRUE ~ suppressWarnings(parse_date_time(Date_Time, orders = c("mdy HMS", "mdy HM", "ymd HMS", "ymd HM")))
    )) %>%
    mutate(Date_Time = as.POSIXct(Date_Time, tz = "UTC", format = "%Y-%m-%d %H:%M:%S"))
} else {
  warning("Column 'Date_Time' not found in species_data.")
}
  
  # Remove unwanted species
species_data <- species_data %>%
  filter(!Species %in% c("Mice, voles, shrews", "Cougar", "Bird", "Moose", "Deer", "Martes martes", "Lynx lynx", "NA"))


# Standardize species names to scientific names
species_data$Species <- recode(species_data$Species,
  # Marten species
  "marten" = "Martes sp.",
  "Marten" = "Martes sp.",
  "American Marten" = "Martes americana",
  "Pacific Marten" = "Martes caurina",

  # Bobcat & Lynx
  "bobcat" = "Lynx rufus",
  "Bobcat" = "Lynx rufus",
  "lynx" = "Lynx canadensis",
  "Lynx" = "Lynx canadensis",
  "Lynx canadiensis" = "Lynx canadensis",
  "Canada lynx" = "Lynx canadensis",
  "Canada Lynx" = "Lynx canadensis",
  "Bobcat/Lynx" = "Lynx sp.",

  # Hare & Rabbit
  "hare" = "Lepus americanus",
  "Snowshoe hare" = "Lepus americanus",
  "Snowshoe Hare" = "Lepus americanus",
  "Rabbits and hares" = "Lepus sp.",

  # Wolverine
  "wolverine" = "Gulo gulo",
  "Wolverine" = "Gulo gulo",

  # Skunk
  "skunk" = "Mephitis mephitis",
  "Striped skunk" = "Mephitis mephitis",
  "Striped Skunk" = "Mephitis mephitis",

  # Coyote
  "coyote" = "Canis latrans",
  "Coyote" = "Canis latrans",

  # Weasels
  "weasel" = "Mustela sp.",
  "Mustela sp" = "Mustela sp.",
  "Mustela sp " = "Mustela sp.",
  "Weasels and Ermine" = "Mustela sp.",
  "Long-tailed weasel" = "Neogale frenata",
  "Short-Tailed Weasel" = "Mustela erminea",

  # Fox
  "Red fox" = "Vulpes vulpes",
  "Red Fox" = "Vulpes vulpes",
  "fox" = "Vulpes vulpes",

  # Badger
  "Badger" = "Taxidea taxus",
  "American Badger" = "Taxidea taxus",

  # Fisher
  "Fisher" = "Pekania pennanti",

  # Squirrel
  "Red squirrel" = "Sciurus vulgaris",
  "Red Squirrel" = "Sciurus vulgaris",
  "Red  Squirrel" = "Sciurus vulgaris",
  "Flying Squirrel" = "Glaucomys sabrinus",

  # Marmot
  "Marmot" = "Marmota sp.",
  "Hoary Marmot" = "Marmota sp.",

  # Otter
  "Otter" = "Lontra canadensis",
  "River Otter" = "Lontra canadensis",

  # Other species
  "Raccoon" = "Procyon lotor",
  "Porcupine" = "Erethizon dorsatum"
)

  # Verify all stations in species and deployments match
  unmatched_stations <- species_data %>% filter(!Station_ID %in% deployment_data$Station_ID)
  
  # Store results
  master_data[[project_name]] <- list(deployment = deployment_data, species = species_data, unmatched = unmatched_stations)
}

# Combine all deployments and species data into master dataframes
master_deployment <- bind_rows(lapply(master_data, "[[", "deployment"))
master_species <- bind_rows(lapply(master_data, "[[", "species"))

``` 


Load the SPI camera data files and standardize to the same formatting as above.


```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(lubridate)

# File paths
file.list <- list.files(path = "C:/LocalR/Merge_datacollation_2025/SPI_cameradata_March2025", 
                        pattern = "*.xlsx", full.names = TRUE)
master_spi_data <- list()  # Initialize storage list

# Column mapping
column_mapping <- c(
  "PROJECT_NAME" = "Project_name",
  "PROJECT_ID" = "SPI_Project_ID",
  "DESIGN_COMPONENT_LABEL" = "Station_ID",
  "LATITUDE" = "Latitude",
  "LONGITUDE" = "Longitude",
  "VISIT_DATE" = "Start_Deployment_date",
  "VISIT_END_DATE" = "End_Deployment_date",
  "TRAP_MODEL" = "Camera_model",
  "TRAP_MAKE_NAME_CD" = "Camera_make", 
  "BAIT_LURE_TYPE" = "Bait_lure",
  "TRAP_TYPE" = "Target_feature",
  "COUNT" = "Count",
  "SCIENTIFIC_NAME" = "Species",
  "OBSERVATION_DATETIME" = "Date_Time",
  "TRAP_HEIGHT" = "Camera_height",
  "VISIT_TIME_SPAN" = "Camera_days"
)

# Function to standardize column names
standardize_columns <- function(df) {
  colnames(df) <- recode(colnames(df), !!!column_mapping)  
  return(df)
}

# Loop through each file
for (i in seq_along(file.list)) {
  file <- file.list[i]
  
  # Extract project name from filename
  file_name <- gsub("^.*data_(.*?)\\-.*$", "\\1", basename(file))
  
  # Read data and standardize column names
  spi_cam <- read_excel(file, sheet = 1) %>% standardize_columns()
  spi_sp <- read_excel(file, sheet = 2) %>% standardize_columns()
  
  # Standardize deployment data
  deployment_spi_data <- spi_cam %>%
    select(any_of(c("Project_name", "SPI_Project_ID", "Station_ID", "Latitude", "Longitude", "Start_Deployment_date", "End_Deployment_date", "Camera_days", 
                    "Camera_make", "Camera_model", "Camera_height", "Bait_lure", "Target_feature"))) %>%
    mutate(across(c(Start_Deployment_date, End_Deployment_date), ~ suppressWarnings(as.Date(.))))

  # Standardize species data
  species_spi_data <- spi_sp %>%
    select(any_of(c("Project_name", "SPI_Project_ID", "Station_ID", "Species", "Count", "Date_Time"))) %>%
    mutate(Date_Time = parse_date_time(Date_Time, orders = c("ymd HMS", "ymd HM", "ymd H", "ymd")))

  # Verify unmatched stations
  unmatched_spi_stations <- species_spi_data %>% filter(!Station_ID %in% deployment_spi_data$Station_ID)
  
  # Store results
  master_spi_data[[file_name]] <- list(deployment = deployment_spi_data, 
                                       species = species_spi_data, 
                                       unmatched = unmatched_spi_stations)
}

# Combine all deployments and species data into master dataframes
master_spi_deployment <- bind_rows(lapply(master_spi_data, "[[", "deployment"))
master_spi_species <- bind_rows(lapply(master_spi_data, "[[", "species"))

```

Merge camera data from shiny app projects, SPI, FN and other government secure files and
rarify the species detection data to 30 minute independent events
```{r}
## Join all deployments
# Ensure all columns in master_spi_deployment are present in master_deployment
missing_cols <- setdiff(names(master_spi_deployment), names(master_deployment))
master_deployment[missing_cols] <- NA

# Bind rows, keeping all columns from master_spi_deployment
all_deployments <- bind_rows(master_spi_deployment, master_deployment)

# View result
head(all_cam_deployments)

#Join all species detections
# Ensure all columns in master_spi_deployment are present in master_deployment
missing_cols <- setdiff(names(master_spi_species), names(master_species))
master_species[missing_cols] <- NA

# Bind rows, keeping all columns from master_spi_species
all_species_cams <- bind_rows(master_spi_species, master_species)

# View result
head(all_species_cams)

## Save a copy to date:
setwd("C:/LocalR/Merge_datacollation_2025/Output_camera_detections") 
write.csv(all_cam_deployments, file="All_cam_deployments_March2025.csv")
write.csv(all_species_cams, file="All_camsp_detections_March2025.csv")

```

Join to project metadata file 
  - Data source (Academic, Government-Not in SPI, First Nations, SPI)
  - Project ID (if in SPI)
  - Contact information: First Name, Last Name, E-mail
  - Project Data Sharing Permissions
  - Comments
```{r}

```


Create a basic interactive map of the camera deployments

```{r}
library(tmap)
library(sf)
library(dplyr)
library(viridis)
library(tmap)

# Convert master_deployment to an sf object, ensuring valid coordinates
camera_stations <- master_deployment %>%
  drop_na(Latitude, Longitude, Project_name) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)  # WGS84 CRS

# Create a numeric factor for Project_name
camera_stations$Project_ID <- as.numeric(as.factor(camera_stations$Project_name))

# Set tmap mode to interactive
tmap_mode("view")

# Use a continuous color scale instead
tm_basemap("OpenStreetMap") +
  tm_shape(camera_stations) +
  tm_dots(col = "Project_ID", palette = "viridis", 
          size = 0.6, alpha = 0.9, title = "Project (by ID)") +
  tm_layout(title = "Camera Station Locations by Project",
            legend.outside = TRUE)


```

Create a basic interactive map from the camera species detections

```{r}
# Load required libraries
library(ggplot2)
library(ggmap)
library(dplyr)
# install.packages("tmap")
library(tmap)
# install.packages("viridis")
library(viridis)


# Ensure Station_ID is the same type in both dataframes
master_species_join <- master_species %>%
  left_join(
    master_deployment %>%
      select(Station_ID, Latitude, Longitude) %>%
      distinct(Station_ID, .keep_all = TRUE),  # Keep unique Station_ID-Latitude-Longitude
    by = "Station_ID"
  )

# Filter for Pekania pennanti and remove NA coordinates
species_map_data <- master_species_join %>%
  filter(Species == "Pekania pennanti") %>%
  drop_na(Latitude, Longitude) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)  # Convert to sf spatial object

# Interactive map with OpenStreetMap
tmap_mode("view")  # Use "plot" instead for a static map
tm_basemap("OpenStreetMap") +
  tm_shape(species_map_data) +
  tm_dots(col = "red", size = 0.2) +
  tm_layout(title = "Fisher Camera Detections in British Columbia")
```

Extract species specific data observations as its own csv file with latitude and longitude associated 

```{r}
# Extract coordinates
species_map_data <- species_map_data %>%
  mutate(Latitude = st_coordinates(.)[, 2],  # Extract Y (Latitude)
         Longitude = st_coordinates(.)[, 1]) %>%  # Extract X (Longitude)
  st_drop_geometry()  # Remove geometry column

# View the new dataframe
head(species_map_data)

## Export
setwd("C:/LocalR/Merge_datacollation_2025/Output_camera_detections")
write.csv(species_map_data, file="Fisher_cam_detections_temp_March2025.csv")

fisher_detection_data <- read.csv("Fisher_cam_detections_temp_March2025.csv")
# Filter for Pekania pennanti and remove NA coordinates
fisher_map_sf <- fisher_detection_data %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)  # Convert to sf spatial object

# Interactive map with OpenStreetMap
tmap_mode("view")  # Use "plot" instead for a static map
tm_basemap("OpenStreetMap") +
  tm_shape(fisher_map_sf) +
  tm_dots(col = "red", size = 0.2) +
  tm_layout(title = "Fisher Camera Detections in British Columbia")
```





Let's map Marten camera detections
```{r}
# Define custom colors
custom_colors <- c("Martes caurina" = "#00796B",  # Dark Teal
                   "Martes sp." = "#D84315",      # Dark Orange
                   "Martes americana" = "#4A148C")  # Dark Purple

# Filter for Marten species and remove NA coordinates
species_map_data <- master_species_join %>%
  filter(Species %in% c("Martes caurina", "Martes sp.", "Martes americana")) %>%
  drop_na(Latitude, Longitude) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)  # Convert to sf spatial object

# Interactive map with OpenStreetMap
tmap_mode("view")  # Use "plot" instead for a static map
tm_basemap("OpenStreetMap") +
  tm_shape(species_map_data) +
  tm_dots(col = "Species", palette = custom_colors, 
          size = 0.4, title = "Species") +  # Increased point size
  tm_layout(title = "Marten Camera Detections in British Columbia",
            legend.outside = TRUE)  # Move legend outside for better visibility


```



Map red fox camera detections out of interest
```{r}

# Filter for Pekania pennanti and remove NA coordinates
species_map_data <- master_species_join %>%
  filter(Species == "Vulpes vulpes") %>%
  drop_na(Latitude, Longitude) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)  # Convert to sf spatial object

# Interactive map with OpenStreetMap
tmap_mode("view")  # Use "plot" instead for a static map
tm_basemap("OpenStreetMap") +
  tm_shape(species_map_data) +
  tm_dots(col = "red", size = 0.2) +
  tm_layout(title = "Red Fox Camera Detections in British Columbia")
```
```{r}

```

