---
title: "Fisher_VHF_QAQC"
output: html_document
date: "2025-04-24"
By Siobhan Darlington
---

The objectives of this script are to:

1) Compile and clean VHF triangulated fisher data for the Columbia population in one dataframe, incorporating the following data sources: File maker files (8 total), BC Wild ground telemetry, BC Wild Aerial telemetry, Live captures, Trapper data, Den tree data, Clappia habitat data, and Aerial data from 2023. 
2) Verify a subset of triangulated bearings to check for erroneous fixes.
3) Clean and finalize a complete VHF dataset
4) Create KDE home ranges for fisher with sufficient data, mapping their VHF locations and HRs, and den tree locations. 

 Code chunk directory:
 
 1. Load libraries
 2. Combine filemaker locations, remove duplicates and keep element locations over triangulated ones
 3. Alternatively, work with Rich's compiled filemaker data and retain UTMs and error_area whenever possible. I recommend adding the lat/longs from step 2. to this sheet due to location errors present. 
4. Check the filemaker compiled bearings and flag potential errors
5. Get UTMs and Flag distances that are too far apart and map
6. Map triangulated locations vs. the bearings 
7. Upload Live capture, trapping, and Aerial 2023 sheet Rich compiled
8. Upload ground telemetry from BC Wild and standardize the column names
9. Extract BC wild ground telemetry bearings from string
10. Overwrite UTMs with conversion from correct Lat/Longs, Remove Precision = "discard"
11. Map flagged bearings in BC wild
12. Upload ground telemetry from Clappia 
13. Upload BC Wild Aerial telemetry and standardize column names
14. Combine all datasets, extract the following columns:
15. Add Shannon's denning data by column bind, and by row bind for new data
16. Fix UTMs for all data, Remove additional duplicates by individual, datetime, and location
17. Precision category cleaning, removing outliers and deletemes
18. Check sample sizes by Fisher_ID and Radiolocation_type, export
19. Map the data to look for any outliers visually
20. Run a prelim AKDE on the fisher locations and map with the den trees in red, export html map
21. Export final data and home range shapefiles

1. Load libraries
```{r}
library(dplyr)
library(geosphere)
library(lubridate)
library(leaflet)
library(sf)
library(htmlwidgets)
library(purrr)
library(RColorBrewer)
library(tibble)
library(tidyr)
library(jsonlite)
library(stringr)
library(leaflet)
library(viridis)
```

2. Combine filemaker locations, remove duplicates and keep element locations over triangulated ones

```{r}

getwd()

badger1.n <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker Badger1 new-2023.csv")
badger1.o <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker Badger1 old-2023.csv")
fisher1.n <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker Fisher1 new-2023.csv")
fisher1.o <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker Fisher1 old-2023.csv")
fisher2 <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker Fisher2-2023.csv")
rf <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker RF-2023.csv")
rw.o <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker RW old-2023.csv")
rw.n <- read.csv(file="Input_VHF_data_sheets/Ground telemetry File maker RW new-2023.csv")

### For all of these above sheets, I took Autocoord_temp and made it three columns "Latitude, Longitude, Error_GPS"
## Note that other field exist such as Autocoord_temp.2 and Autocoord_saved
## I also corrected Errror_area to Error_area because it was annoying me. 

badger1.n$Data_source <- "badger1_newtemp"
fisher1.n$Data_source <- "fisher1_newtemp"
badger1.o$Data_source <- "badger1_oldtemp"
fisher1.o$Data_source <- "fisher1_oldtemp"
fisher2$Data_source <- "fisher2_oldtemp"
rf$Data_source <- "RF_newtemp"
rw.o$Data_source <- "RW_oldtemp"
rw.n$Data_source <- "RW_newtemp"

## add missing columns to fisher2 and rf
fisher2$Mark_for_deletion <- ""
fisher2$FPP_completed <- ""
fisher2$Selected_for_FPP <- ""
fisher2$PCP_completed <- ""

fisher1.o$Mark_for_deletion <- ""
fisher1.o$FPP_completed <- ""
fisher1.o$Selected_for_FPP <- ""
fisher1.o$PCP_completed <- ""

badger1.o$Mark_for_deletion <- ""
badger1.o$FPP_completed <- ""
badger1.o$Selected_for_FPP <- ""
badger1.o$PCP_completed <- ""

rw.o$Mark_for_deletion <- ""
rw.o$FPP_completed <- ""
rw.o$Selected_for_FPP <- ""
rw.o$PCP_completed <- ""

rf$Mark_for_deletion <- ""
rf$FPP_completed <- ""
rf$Selected_for_FPP <- ""
rf$PCP_completed <- ""

setdiff(names(badger1.o), names(rw.n))  # Columns in badger1.o but not in badger1.n
setdiff(names(rw.n), names(fisher2))  # Columns in badger1.n but not in badger1.o


fm_data <- rbind(badger1.n, badger1.o, fisher1.n, fisher1.o, fisher2, rf, rw.o , rw.n)

## Convert the "Best_UTM_easting" and "Best_UTM_northing" and put the values in "Actual_latitude" and "Actual_longitude"

# Make a copy to preserve original
# fm_data_with_coords <- fm_data %>%
#   filter(!is.na(Best_UTM_easting) & !is.na(Best_UTM_northing))
# 
# # Convert to sf using UTM zone 10N (EPSG:32701)
# fm_data_sf <- st_as_sf(fm_data_with_coords, coords = c("Best_UTM_easting", "Best_UTM_northing"), crs = 332710, remove = FALSE)
# 
# # Transform to WGS84 (EPSG:4327)
# fm_data_sf <- st_transform(fm_data_sf, crs = 4327)
# 
# # Extract decimal degrees into separate columns
# coords <- st_coordinates(fm_data_sf)
# 
#  #Create a logical index for rows with valid UTM coordinates
# valid_utm <- !is.na(fm_data$Best_UTM_easting) & !is.na(fm_data$Best_UTM_northing)
# 
# # Create a logical index for where Actual_longitude is NA (within the subset)
# needs_update <- is.na(fm_data$Actual_longitude[valid_utm])
# 
# # Update only those rows in the original data
# fm_data$Actual_longitude[valid_utm][needs_update] <- coords[needs_update, "X"]
# fm_data$Actual_latitude[valid_utm][needs_update] <- coords[needs_update, "Y"]

### Replace latitude and longitude if Actual_latitude and Actual_longitude are present

# Replace Latitude with Actual_latitude where not NA
fm_data$Latitude[!is.na(fm_data$Actual_latitude)] <- fm_data$Actual_latitude[!is.na(fm_data$Actual_latitude)]

# Replace Longitude with Actual_longitude where not NA
fm_data$Longitude[!is.na(fm_data$Actual_longitude)] <- fm_data$Actual_longitude[!is.na(fm_data$Actual_longitude)]

## Create a UTM_Easting and UTM_Northing column from the new Latitude and Longitude columns under UTM 11
# Step 1: Identify rows with valid coordinates
valid_coords <- !is.na(fm_data$Latitude) & !is.na(fm_data$Longitude)

# Step 2: Create a temporary sf object for valid rows, using UTM Zone 10N (EPSG:32710)
fm_data_utm <- fm_data[valid_coords, ] %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4327, remove = FALSE) %>%
  st_transform(crs = 32710)  # ← UTM Zone 10N

# Step 3: Extract UTM coordinates
utm_coords <- st_coordinates(fm_data_utm)

# Step 4: Create new columns and insert UTM values
fm_data$UTM_Easting <- NA_real_
fm_data$UTM_Northing <- NA_real_

fm_data$UTM_Easting[valid_coords] <- utm_coords[, 1]
fm_data$UTM_Northing[valid_coords] <- utm_coords[, 2]

## now, let's remove duplicates using CreationTimestamp, Fisher_ID, Latitude and Longitude while keeping the Radiolocation_type marked as "Element"

# Remove duplicates with preference for 'Element' in Radiolocation_type
fm_data_cleaned <- fm_data %>%
  # Arrange to prioritize 'Element' entries
  arrange(Fisher_ID, CreationTimestamp, 
          desc(Radiolocation_type == "Element")) %>%
  # Remove duplicates based on specified columns
  distinct(Fisher_ID, CreationTimestamp, .keep_all = TRUE)

## Make a new precision column, replace NA values in Precision_category with Home Range level precision (4 or 5)

fm_data_cleaned <- fm_data_cleaned %>%
  mutate(Precision_adjusted = if_else(is.na(Precision_category) | Precision_category == "", 
                                      "4", 
                                      as.character(Precision_category)))


write.csv(fm_data_cleaned, file="Ground_Telemetry_cleaned_Filemaker_May22025.csv")
```

3. Alternatively, work with Rich's compiled filemaker data and retain UTMs and error_area whenever possible. I recommend adding the lat/longs from step 2. to this sheet due to location errors present. 

```{r}

fm_compile <- read.csv(file="Input_VHF_data_sheets/FULL_compiled_240521_locations_exported.csv", header=T)

fm_compile1 <- fm_compile %>% 
  mutate(UTM_easting = if_else(!is.na(Best_UTM_easting),Best_UTM_easting,
                               if_else(!is.na(Actual_UTM_easting), Actual_UTM_easting,
                                       if_else(!is.na(Flagging_UTM_easting), Flagging_UTM_easting,
                                               Triang_easting))),
         UTM_northing = if_else(!is.na(Best_UTM_northing),Best_UTM_northing,
                               if_else(!is.na(Actual_UTM_northing), Actual_UTM_northing,
                                       if_else(!is.na(Flagging_UTM_northing), Flagging_UTM_northing,
                                               Triang_northing))))


## Ok now need to fill in the UTM gaps  - I think we can pull from dataframe above.

# Step 1: Join only the needed columns
fm_compile1 <- fm_compile1 %>%
  left_join(fm_data_cleaned %>%
             dplyr::select(Location_UUID, UTM_Easting, UTM_Northing, Data_source),
            by = "Location_UUID")

# Step 2: Fill in NA values in UTM_easting/northing from the new columns
fm_compile1 <- fm_compile1 %>%
  mutate(
    UTM_easting = if_else(is.na(UTM_easting), UTM_Easting, UTM_easting),
    UTM_northing = if_else(is.na(UTM_northing), UTM_Northing, UTM_northing)
  )

## Make new precision category as above

fm_compile1 <- fm_compile1 %>%
  mutate(Precision_adjusted = if_else(is.na(Precision_category) | Precision_category == "", 
                                      "4", 
                                      as.character(Precision_category)))

## Remove duplicates as above noting that RW already removed many duplicates but just to check

# Remove duplicates with preference for 'Element' in Radiolocation_type
fm_compile1 <- fm_compile1 %>%
  # Arrange to prioritize 'Element' entries
  arrange(Fisher_ID, CreationTimestamp, 
          desc(Radiolocation_type == "Element")) %>%
  # Remove duplicates based on specified columns
  distinct(Fisher_ID, CreationTimestamp, .keep_all = TRUE) ## removed 1 obs

fm_compile1<- fm_compile1 %>%
  rename(
    Error_area = Error_area,
    Error_easting = Triang_easting,
    Error_northing = Triang_northing
  )

## Add Latitude and Longitude to this dataframe from UTM_easting and UTM_northing
# Convert fm_compile1 to an sf object using UTM Zone 10N (EPSG:32610)
fm_compile1_sf <- fm_compile1 %>%
  st_as_sf(coords = c("UTM_easting", "UTM_northing"), crs = 32610)

# Transform to WGS84 (EPSG:4326) to get lat/lon
fm_compile1_latlon <- fm_compile1_sf %>%
  st_transform(crs = 4326)

# Extract the lat/lon coordinates back into columns
fm_compile1 <- fm_compile1_latlon %>%
  mutate(
    Longitude = st_coordinates(.)[, 1],
    Latitude = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry() 

table(fm_compile1$Precision_adjusted)

fm_compile1 <- subset(fm_compile1, fm_compile1$Precision_adjusted!="discard")

write.csv(fm_compile1, file="FM_Compile_RW_SD_May2025.csv")
#fm_compile1 <- read.csv(file="Input_VHF_data_sheets/FM_Compile_RW_SD_May2025.csv", header=T)
```


4. Check the filemaker compiled bearings and flag potential errors
```{r}
##Loading a file of bearing from Rory to check the quality of the data, the bearings from FM are all in separate sheets...

rf_bearings <- read.csv(file="Input_VHF_data_sheets/RF_Bearings_Filemaker_Oct2023.csv", header=T)

## Replace missing coordinates for lat/long

# Create an sf object from UTM coordinates (Zone 10N, EPSG:32610)
utm_sf <- rf_bearings %>%
  filter(is.na(Bearing_latitude) | is.na(Bearing_longitude)) %>%
  st_as_sf(coords = c("UTM_easting", "UTM_northing"), crs = 32610, remove = FALSE)

# Transform to WGS84 (lat/lon)
wgs84_sf <- st_transform(utm_sf, crs = 4326)

# Extract the lat/lon values and update only the NA rows in the original dataframe
rf_bearings <- rf_bearings %>%
  mutate(
    Bearing_longitude = ifelse(is.na(Bearing_longitude),
                                st_coordinates(wgs84_sf)[, 1], 
                                Bearing_longitude),
    Bearing_latitude = ifelse(is.na(Bearing_latitude),
                               st_coordinates(wgs84_sf)[, 2], 
                               Bearing_latitude)
  )

## I need to attach the bearing data to the file maker data to obtain the triangulated values for flagging

# Step 1: Assign bearing numbers within each Location_UUID
rf_bearings_numbered <- rf_bearings %>%
  group_by(Location_UUID) %>%
  mutate(Bearing_number = row_number()) %>%
  ungroup()

# Step 2: Pivot wider to one row per Location_UUID
rf_bearings_wide <- rf_bearings_numbered %>%
  dplyr::select(Location_UUID, Bearing_number, Bearing, Bearing_latitude, Bearing_longitude) %>%
  pivot_wider(
    names_from = Bearing_number,
    values_from = c(Bearing, Bearing_latitude, Bearing_longitude),
    names_glue = "Bearing{Bearing_number}_{.value}"
  ) %>%
 rename_with(~ gsub("_Bearing$", "", .x)) %>%  # Change BearingN_Bearing -> BearingN
  rename_with(~ gsub("_Bearing_latitude$", "_latitude", .x)) %>%  # Change BearingN_Bearing_latitude -> BearingN_latitude
  rename_with(~ gsub("_Bearing_longitude$", "_longitude", .x))  


# Step 3: Join to fm_compile1
fm_compile1_joined <- fm_compile1 %>%
  left_join(rf_bearings_wide, by = "Location_UUID")
```


5. Get UTMs and Flag distances that are too far apart and map

```{r}
# Function to convert lat/long to UTM (Zone 10)
convert_to_utm <- function(lat, long, zone = 10) {
  if (is.na(lat) || is.na(long)) return(data.frame(Easting = NA, Northing = NA))
  
  point <- st_sfc(st_point(c(as.numeric(long), as.numeric(lat))), crs = 4326)
  point_utm <- st_transform(point, crs = paste0("+proj=utm +zone=", zone, " +datum=WGS84 +units=m +no_defs"))
  coords <- st_coordinates(point_utm)
  data.frame(Easting = coords[1], Northing = coords[2])
}

# Loop to convert bearing coordinates to UTM and calculate distance
for (i in 1:9) {
  lat_col <- paste0("Bearing", i, "_latitude")
  long_col <- paste0("Bearing", i, "_longitude")
  
  utm_e_col <- paste0("Bearing", i, "_easting")
  utm_n_col <- paste0("Bearing", i, "_northing")
  dist_col <- paste0("Bearing", i, "_distance_m")
  
  utm_coords <- mapply(convert_to_utm, fm_compile1_joined[[lat_col]], fm_compile1_joined[[long_col]], SIMPLIFY = FALSE)
  utm_df <- do.call(rbind, utm_coords)
  
  fm_compile1_joined[[utm_e_col]] <- utm_df$Easting
  fm_compile1_joined[[utm_n_col]] <- utm_df$Northing
  
  # Compute Euclidean distance from triangulated UTM (using correct UTM_Easting / UTM_Northing)
  fm_compile1_joined[[dist_col]] <- with(fm_compile1_joined, ifelse(
    is.na(utm_df$Easting) | is.na(UTM_Easting), NA,
    sqrt((utm_df$Easting - UTM_Easting)^2 + (utm_df$Northing - UTM_Northing)^2)
  ))
}

# Add flag if any distance exceeds 3000m
fm_compile1_joined <- fm_compile1_joined %>%
  mutate(
    Bearing_flagged_3km = if_else(
      pmax(
        Bearing1_distance_m,
        Bearing2_distance_m,
        Bearing3_distance_m,
        Bearing4_distance_m,
        Bearing5_distance_m,
        Bearing6_distance_m,
        Bearing7_distance_m,
        Bearing8_distance_m,
        Bearing9_distance_m,
        na.rm = TRUE
      ) > 3000,
      "flag",
      NA_character_
    )
  )

# Add which specific bearings are >3000m
fm_compile1_joined <- fm_compile1_joined %>%
  mutate(
    Flagged_bearing_3km = pmap_chr(
      list(Bearing1_distance_m,
           Bearing2_distance_m,
           Bearing3_distance_m,
           Bearing4_distance_m,
           Bearing5_distance_m,
           Bearing6_distance_m,
           Bearing7_distance_m,
           Bearing8_distance_m,
           Bearing9_distance_m),
      function(d1, d2, d3, d4, d5, d6, d7, d8, d9) {
        bearings <- c()
        if (!is.na(d1) && d1 > 3000) bearings <- c(bearings, "Bearing1")
        if (!is.na(d2) && d2 > 3000) bearings <- c(bearings, "Bearing2")
        if (!is.na(d3) && d3 > 3000) bearings <- c(bearings, "Bearing3")
        if (!is.na(d4) && d4 > 3000) bearings <- c(bearings, "Bearing4")
        if (!is.na(d5) && d5 > 3000) bearings <- c(bearings, "Bearing5")
        if (!is.na(d6) && d6 > 3000) bearings <- c(bearings, "Bearing6")
        if (!is.na(d7) && d7 > 3000) bearings <- c(bearings, "Bearing7")
        if (!is.na(d8) && d8 > 3000) bearings <- c(bearings, "Bearing8")
        if (!is.na(d9) && d9 > 3000) bearings <- c(bearings, "Bearing9")
        if (length(bearings) == 0) NA_character_ else paste(bearings, collapse = ", ")
      }
    )
  )


```


6. Map triangulated locations vs. the bearings 

```{r}
# Add a row ID and bearing flag column
fm_compile1_joined <- fm_compile1_joined %>%
  mutate(row_id = row_number(),
         Bearing_flagged = ifelse(is.na(Bearing_flagged_3km), "none", Bearing_flagged_3km))

# Ensure coordinates are numeric
fm_compile1_joined <- fm_compile1_joined %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    across(matches("Bearing[1-9]_(latitude|longitude|bearing)"), as.numeric)
  )

# Filter rows with at least one valid bearing coordinate
fm_compile1_filtered <- fm_compile1_joined %>%
  filter(
    rowSums(
      across(matches("Bearing[1-9]_(latitude|longitude)"), ~ !is.na(.))
    ) > 0
  )

# Create Fisher_ID color palette
fisher_ids <- unique(fm_compile1_filtered$Fisher_ID)
viridis_colors <- c(
  "#440154", "#482878", "#3E4A89", "#31688E", "#26828E", "#1F9E89",
  "#35B779", "#6DCD59", "#B4DE2C", "#FDE725", "#F9F9F9"
)
fisher_palette <- colorFactor(palette = viridis_colors, domain = fisher_ids)

# Build map
map <- leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%

  # Triangulated (non-flagged)
  addCircleMarkers(
    data = fm_compile1_filtered %>% filter(Bearing_flagged != "flag"),
    lng = ~Longitude, lat = ~Latitude,
    color = ~fisher_palette(Fisher_ID),
    radius = 5,
    popup = ~paste("RowID:", row_id,
                   "<br>Fisher_ID:", Fisher_ID,
                   "<br>Triangulated:", Latitude, ",", Longitude),
    group = "Triangulated (By Fisher_ID)"
  ) %>%

  # Flagged triangulated
  addCircleMarkers(
    data = fm_compile1_filtered %>% filter(Bearing_flagged == "flag"),
    lng = ~Longitude, lat = ~Latitude,
    color = "red",
    radius = 5,
    popup = ~paste("RowID:", row_id,
                   "<br>Fisher_ID:", Fisher_ID,
                   "<br>FLAGGED:", Latitude, ",", Longitude),
    group = "Flagged (Red)"
  )

for (i in 1:9) {
  lat_col <- paste0("Bearing", i, "_latitude")
  lon_col <- paste0("Bearing", i, "_longitude")
  bearing_col <- paste0("Bearing", i)  # <- changed this line

  # Flagged bearings
  flagged_bearings <- fm_compile1_filtered %>%
    filter(Bearing_flagged == "flag", !is.na(.data[[lat_col]]) & !is.na(.data[[lon_col]])) %>%
    mutate(lat = .data[[lat_col]], lon = .data[[lon_col]], bearing_val = .data[[bearing_col]])

  map <- map %>%
    addCircleMarkers(
      data = flagged_bearings,
      lng = ~lon, lat = ~lat,
      color = "gray",
      radius = 5,
      popup = ~paste0("RowID: ", row_id,
                      "<br>Fisher_ID: ", Fisher_ID,
                      "<br>FLAGGED Bearing", i,
                      "<br>Location: ", round(lat, 5), ", ", round(lon, 5),
                      "<br>Bearing: ", bearing_val),
      group = "Flagged Bearings (Gray)"
    )

  # Regular bearings
  regular_bearings <- fm_compile1_filtered %>%
    filter(Bearing_flagged != "flag", !is.na(.data[[lat_col]]) & !is.na(.data[[lon_col]])) %>%
    mutate(lat = .data[[lat_col]], lon = .data[[lon_col]], bearing_val = .data[[bearing_col]])

  map <- map %>%
    addCircleMarkers(
      data = regular_bearings,
      lng = ~lon, lat = ~lat,
      color = "black",
      radius = 5,
      popup = ~paste0("RowID: ", row_id,
                      "<br>Fisher_ID: ", Fisher_ID,
                      "<br>Bearing", i,
                      "<br>Location: ", round(lat, 5), ", ", round(lon, 5),
                      "<br>Bearing: ", bearing_val),
      group = "Bearings (Black)"
    )
}

# Add legend and controls
map <- map %>%
  addLegend(
    position = "bottomright",
    pal = fisher_palette,
    values = fm_compile1_filtered$Fisher_ID,
    title = "Fisher_ID",
    opacity = 2
  ) %>%
  addLayersControl(
    overlayGroups = c("Triangulated (By Fisher_ID)", "Flagged (Red)", "Bearings (Black)", "Flagged Bearings (Gray)"),
    options = layersControlOptions(collapsed = FALSE)
  )

# Show map
map

# Save map as HTML
saveWidget(map, "fm_compile1_map_with_rowid.html", selfcontained = TRUE)

# Drop list-type columns
fm_compile1_flat <- fm_compile1_joined %>%
 dplyr::select(where(~ class(.) != "list"))


### Note that there are many bearings showing up at Larry's house and only 2 are coming up a "flagged" when they all should be. # Will remove any "discard" locations and run the map again. 

saveWidget(map, file = "html_maps/Filemaker_bearings_map.html", selfcontained = TRUE)

```


7. Upload Live capture, trapping, and Aerial 2023 sheet Rich compiled

```{r}
aerial_caps <- read.csv(file="Input_VHF_data_sheets/Aerial_Capture_Trap_2023.csv", header=T)

aerial_caps <- aerial_caps %>%
  rename(Data_source = Source)

aerial_caps$Precision_adjusted <- aerial_caps$Precision_category
aerial_caps$Error_area <- ""
aerial_caps$Error_northing <- ""
aerial_caps$Error_easting <- ""

# Convert to sf object with UTM Zone 10N (EPSG:32610)
aerial_caps_sf <- st_as_sf(aerial_caps, coords = c("UTM_easting", "UTM_northing"), crs = 32610)

# Transform to WGS84 (EPSG:4326)
aerial_caps_sf <- st_transform(aerial_caps_sf, crs = 4326)

# Extract coordinates and add them to the original dataframe
aerial_caps <- aerial_caps %>%
  bind_cols(
    st_coordinates(aerial_caps_sf) %>%
      as.data.frame() %>%
      rename(Longitude = X, Latitude = Y)
  )

```

8. Upload ground telemetry from BC Wild and standardize the column names

```{r}
bcw_ground <- read.csv(file="Input_VHF_data_sheets/Ground telemetry BC Wild-02may2025.csv")

bcw_ground$Radiolocation_type <- ifelse(bcw_ground$element_identified=="yes", "Element", "Triangulation")

# Rename specific columns
bcw_ground <- bcw_ground %>%
  rename(
    Latitude = latitude,
    Longitude = longitude,
    Fisher_ID = animal_id,
    UTM_easting = easting, 
    UTM_northing = northing,
    Error_easting = easting_error,
    Error_northing = northing_error,
    Error_area = error_area,
    Date = date,
    Location_UUID = location_id,
  )

# Format the datetime to "yyyy-mm-dd HH:mm:ss"
bcw_ground$Date<- as.Date(bcw_ground$Date, "%Y-%m-%d")

## Add data source
bcw_ground$Data_source <- "BCWild Ground Telemetry"

## Calculate Precision_category and add Precision_adjusted

bcw_ground <- bcw_ground %>%
  mutate(
    Precision_category = case_when(
      Radiolocation_type == "Element" ~ "0",
      Error_area == 0 ~ "0",
      Error_area > 0 & Error_area <= 3000 ~ "1",
      Error_area > 3000 & Error_area <= 8000 ~ "2",
      Error_area > 8000 & Error_area <= 30000 ~ "3",
      Error_area > 30000 & Error_area <= 200000 ~ "4",
      Error_area > 200000 & Error_area <= 750000 ~ "5",
      Error_area > 750000 ~ "discard",
      TRUE ~ NA_character_
    )
  )


bcw_ground <- bcw_ground %>%
  mutate(Precision_adjusted = if_else(is.na(Precision_category) | Precision_category == "", 
                                      "4", 
                                      as.character(Precision_category)))


## Remove duplicates

# Remove duplicates with preference for 'Element' in Radiolocation_type
bcw_ground_cleaned <- bcw_ground %>%
  # Arrange to prioritize 'Element' entries
  arrange(Fisher_ID, Date,
          desc(Radiolocation_type == "Element")) %>%
  # Remove duplicates based on specified columns
  distinct(Fisher_ID, Date, .keep_all = TRUE) ## dropped 2 locations


## write the cleaned version
write.csv(bcw_ground_cleaned, file="BCWild_Ground_cleaned_May22025.csv")
```

9. Extract BC wild ground telemetry bearings from string

OK now - we are going to extract all of the bearings from BC Wild ground telemetry in the painful string format that they gave us, and convert them to UTMs and measure the distance between these bearings and the projected coordinate to flag problem locations. I will also code this to map bearings in one colour and their triangulated value in another.
```{r}
## Extract the bearings from this sheet
bcw_ground1 <- bcw_ground_cleaned

# Replace "NULL" with an empty string in the triangulation column
bcw_ground1 <- bcw_ground1 %>%
  mutate(triangulation = ifelse(triangulation == "NULL", "", triangulation))

# Clean up the triangulation strings by removing escape characters
bcw_ground1$cleaned_triangulation <- str_replace_all(bcw_ground1$triangulation, "\\\\", "")

# Extract all latitudes and longitudes using regular expressions
bcw_ground1 <- bcw_ground1 %>%
  mutate(
    # Extract longitude and latitude values
    coordinates = str_extract_all(cleaned_triangulation, '"longitude":([0-9.-]+),"latitude":([0-9.-]+)'),
    coordinates = lapply(coordinates, function(x) str_extract_all(x, "[0-9.-]+"))
  )

# View the structure of the coordinates column
head(bcw_ground1$coordinates, 5)

library(dplyr)

# Define a function to extract longitude and latitude, handling empty lists
extract_coords <- function(coord_list) {
  if (length(coord_list) > 0 && length(coord_list[[1]]) == 2) {
    return(c(longitude = coord_list[[1]][1], latitude = coord_list[[1]][2]))
  } else {
    return(c(longitude = NA, latitude = NA))  # Return NA if the list is empty
  }
}

# Apply the function to the 'coordinates' column and extract longitude/latitude
bcw_ground1 <- bcw_ground1 %>%
  rowwise() %>%
  mutate(
    coords = list(extract_coords(coordinates)),
    Bearing1_long = coords["longitude"],
    Bearing1_lat = coords["latitude"]
  ) %>%
  ungroup()

# View the result
head(bcw_ground1[, c("Bearing1_lat", "Bearing1_long")], 5)

extract_coordinates <- function(coord_list) {
  # If no coordinates, return 9 rows of NA
  if (length(coord_list) == 0) {
    return(data.frame(
      Longitude = rep(NA, 9),
      Latitude = rep(NA, 9)
    ))
  }
  
  # Extract each coordinate pair, convert to numeric
  coords <- lapply(coord_list, function(x) as.numeric(x))
  
  # Pad with NAs if fewer than 9 points
  while (length(coords) < 9) {
    coords[[length(coords) + 1]] <- c(NA, NA)
  }
  
  # Bind and convert to data frame
  coords_df <- do.call(rbind, coords)
  colnames(coords_df) <- c("Longitude", "Latitude")
  return(coords_df)
}

# Extract coordinates
coords_list <- lapply(bcw_ground1$coordinates, extract_coordinates)

# Now extract manually from the list into columns
bcw_ground1$Bearing1_long <- sapply(coords_list, function(x) x[1, "Longitude"])
bcw_ground1$Bearing1_lat  <- sapply(coords_list, function(x) x[1, "Latitude"])

bcw_ground1$Bearing2_long <- sapply(coords_list, function(x) x[2, "Longitude"])
bcw_ground1$Bearing2_lat  <- sapply(coords_list, function(x) x[2, "Latitude"])

bcw_ground1$Bearing3_long <- sapply(coords_list, function(x) x[3, "Longitude"])
bcw_ground1$Bearing3_lat  <- sapply(coords_list, function(x) x[3, "Latitude"])

bcw_ground1$Bearing4_long <- sapply(coords_list, function(x) x[4, "Longitude"])
bcw_ground1$Bearing4_lat  <- sapply(coords_list, function(x) x[4, "Latitude"])

bcw_ground1$Bearing5_long <- sapply(coords_list, function(x) x[5, "Longitude"])
bcw_ground1$Bearing5_lat  <- sapply(coords_list, function(x) x[5, "Latitude"])

bcw_ground1$Bearing6_long <- sapply(coords_list, function(x) x[6, "Longitude"])
bcw_ground1$Bearing6_lat  <- sapply(coords_list, function(x) x[6, "Latitude"])

bcw_ground1$Bearing7_long <- sapply(coords_list, function(x) x[7, "Longitude"])
bcw_ground1$Bearing7_lat  <- sapply(coords_list, function(x) x[7, "Latitude"])

bcw_ground1$Bearing8_long <- sapply(coords_list, function(x) x[8, "Longitude"])
bcw_ground1$Bearing8_lat  <- sapply(coords_list, function(x) x[8, "Latitude"])

bcw_ground1$Bearing9_long <- sapply(coords_list, function(x) x[9, "Longitude"])
bcw_ground1$Bearing9_lat  <- sapply(coords_list, function(x) x[9, "Latitude"])


# View the result to check
head(bcw_ground1[, c("Bearing1_lat", "Bearing1_long", "Bearing2_lat", "Bearing2_long", "Bearing3_lat", "Bearing3_long", "Bearing4_lat", "Bearing4_long")], 5)

### Now create all bearings in UTM to measure the distance from them to the triangulated version and flag any outside of a 3 km radius. 

# Function to convert lat/long to UTM
convert_to_utm <- function(lat, long, zone = 10) {
  if (is.na(lat) || is.na(long)) return(data.frame(Easting = NA, Northing = NA))
  
  point <- st_sfc(st_point(c(as.numeric(long), as.numeric(lat))), crs = 4326)
  point_utm <- st_transform(point, crs = paste0("+proj=utm +zone=", zone, " +datum=WGS84 +units=m +no_defs"))
  coords <- st_coordinates(point_utm)
  data.frame(Easting = coords[1], Northing = coords[2])
}

# Loop to convert each bearing to UTM and calculate distance
for (i in 1:9) {
  lat_col <- paste0("Bearing", i, "_lat")
  long_col <- paste0("Bearing", i, "_long")
  
  utm_e_col <- paste0("Bearing", i, "_easting")
  utm_n_col <- paste0("Bearing", i, "_northing")
  dist_col <- paste0("Bearing", i, "_distance_m")
  
  utm_coords <- mapply(convert_to_utm, bcw_ground1[[lat_col]], bcw_ground1[[long_col]], SIMPLIFY = FALSE)
  utm_df <- do.call(rbind, utm_coords)
  
  bcw_ground1[[utm_e_col]] <- utm_df$Easting
  bcw_ground1[[utm_n_col]] <- utm_df$Northing
  
  # Calculate Euclidean distance to triangulated point
  bcw_ground1[[dist_col]] <- with(bcw_ground1, ifelse(
    is.na(utm_df$Easting) | is.na(UTM_easting), NA,
    sqrt((utm_df$Easting - UTM_easting)^2 + (utm_df$Northing - UTM_northing)^2)
  ))
}


bcw_ground1 <- bcw_ground1 %>%
  mutate(
    Bearing_flagged_3km = if_else(
      pmax(
        Bearing1_distance_m,
        Bearing2_distance_m,
        Bearing3_distance_m,
        Bearing4_distance_m,
        Bearing5_distance_m,
        Bearing6_distance_m,
        Bearing7_distance_m,
        Bearing8_distance_m,
        Bearing9_distance_m,
        na.rm = TRUE
      ) > 3000,
      "flag",
      NA_character_
    )
  )

## Flag which bearings are off
bcw_ground1 <- bcw_ground1 %>%
  mutate(
    Flagged_bearing_3km = pmap_chr(
      list(Bearing1_distance_m,
        Bearing2_distance_m,
        Bearing3_distance_m,
        Bearing4_distance_m,
        Bearing5_distance_m,
        Bearing6_distance_m,
        Bearing7_distance_m,
        Bearing8_distance_m,
        Bearing9_distance_m),
      function(d1, d2, d3, d4, d5, d6, d7, d8, d9) {
        bearings <- c()
        if (!is.na(d1) && d1 > 3000) bearings <- c(bearings, "Bearing1")
        if (!is.na(d2) && d2 > 3000) bearings <- c(bearings, "Bearing2")
        if (!is.na(d3) && d3 > 3000) bearings <- c(bearings, "Bearing3")
        if (!is.na(d4) && d4 > 3000) bearings <- c(bearings, "Bearing4")
        if (!is.na(d5) && d5 > 3000) bearings <- c(bearings, "Bearing5")
        if (!is.na(d6) && d6 > 3000) bearings <- c(bearings, "Bearing6")
        if (!is.na(d7) && d7 > 3000) bearings <- c(bearings, "Bearing7")
        if (!is.na(d8) && d8 > 3000) bearings <- c(bearings, "Bearing8")
        if (!is.na(d9) && d9 > 3000) bearings <- c(bearings, "Bearing9")
        if (length(bearings) == 0) NA_character_ else paste(bearings, collapse = ", ")
      }
    )
  )

## that gets us a list of possibly incorrect bearings, or at least shows where the triangulated value doesn't line up with bearings.


## Extract the bearings too:
# Extract all bearing values (as character) using regex
bcw_ground1 <- bcw_ground1 %>%
  mutate(
    # Extract all matches of "bearing":"XXX" and pull out the value
    bearing_values = str_extract_all(cleaned_triangulation, '"bearing":"([0-9.]+)"'),
    bearing_values = lapply(bearing_values, function(x) as.numeric(str_extract(x, "[0-9.]+")))
  )

# Function to extract up to 9 bearing values (pad with NA)
extract_bearings <- function(bearing_list) {
  if (length(bearing_list) == 0) {
    return(rep(NA, 9))
  }
  # Pad to length 9
  while (length(bearing_list) < 9) {
    bearing_list <- c(bearing_list, NA)
  }
  return(bearing_list[1:9])
}

# Apply the function to get a 9-column matrix
bearing_matrix <- t(sapply(bcw_ground1$bearing_values, extract_bearings))

# Assign to columns: Bearing1_bearing through Bearing9_bearing
for (i in 1:9) {
  bcw_ground1[[paste0("Bearing", i, "_bearing")]] <- bearing_matrix[, i]
}

```

10. Overwrite UTMs with conversion from correct Lat/Longs, Remove Precision = "discard"

It looks to me like the lat/longs and UTMs are not direct conversions of each other, I'm going to flag UTMs that don't align with the lat/longs in the datasheet. This is because the previous exercise shows me that the bearings for several flagged locations look to be correct - and the map is based on decimal degrees and flags are based on UTMs. 
```{r}
#### it looks to me like some of the UTM values aren't the same as the decimal degrees values in this file? 

# Convert to sf object using WGS84 (EPSG:4326)
bcw_ground1_sf <- st_as_sf(
  bcw_ground1,
  coords = c("Longitude", "Latitude"),
  crs = 4326,
  remove = FALSE
)

# Transform to UTM Zone 10N (EPSG:32610)
bcw_ground1_utm <- st_transform(bcw_ground1_sf, crs = 32610)

# Extract UTM coordinates
utm_coords <- st_coordinates(bcw_ground1_utm)
bcw_ground1$UTME_long <- utm_coords[, "X"]
bcw_ground1$UTMN_lat <- utm_coords[, "Y"]

# Compare with existing UTM columns and flag discrepancies
bcw_ground1 <- bcw_ground1 %>%
  mutate(
    Flag_UTMs = ifelse(
      round(UTME_long) != round(UTM_easting) | round(UTMN_lat) != round(UTM_northing),
      "flag",
      NA
    )
  )

## This produced 1 flagged UTM coordinate

## Let's also get rid of the Precision_adjusted =="discard"

bcw_ground1 <- subset(bcw_ground1, bcw_ground1$Precision_adjusted !="discard")

### There are some errors with denning females
### It looks like 2 F04 entries from Shannon are incorrect, their coordinates are central to F06 territory. 
## April 19 2024 and 2024-06-11 both say it's F04 in her den but the locations are wrong. 
#Date 2024-06-11, -121.6520, 51.66245
#Date 2024-04-19, -121.6520, 51.67340 # suspicious same longitude as the other point, not sure what's up with these coordinates. will remove for now.
# Ensure 'Date' column is of Date type
bcw_ground1 <- bcw_ground1 %>%
  mutate(Date = as.Date(Date))

# Step 1: Remove the specific row
bcw_ground1 <- bcw_ground1 %>%
  filter(!(first_location_id=="F04_24-may-2024" &
           Fisher_ID == "F04"))

# Step 2: Update Fisher_ID for the specified row
bcw_ground1 <- bcw_ground1 %>%
  mutate(Fisher_ID = if_else(
    first_location_id == "F04_19-apr-2024_element" & Fisher_ID == "F04",
    "F06",
    Fisher_ID
  ))


```

11. Map flagged bearings in BC wild
```{r}
# Add a row ID column
bcw_ground1 <- bcw_ground1 %>%
  mutate(row_id = row_number(),
         Bearing_flagged = ifelse(is.na(Bearing_flagged_3km), "none", Bearing_flagged_3km))

# Ensure all coordinates are numeric
bcw_ground1 <- bcw_ground1 %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    across(matches("Bearing[1-9]_(lat|long|bearing)"), as.numeric)
  )

# Keep rows with at least one valid bearing coordinate
bcw_ground1_filtered <- bcw_ground1 %>%
  filter(
    rowSums(
      across(matches("Bearing[1-9]_(lat|long)"), ~ !is.na(.))
    ) > 0
  )


library(RColorBrewer)

# Create color palette for Fisher_IDs
fisher_ids <- unique(bcw_ground1_filtered$Fisher_ID)
# Manually define 11 colors based on Viridis palette
viridis_colors <- c(
  "#440154", # Dark purple
  "#482878", # Purple
  "#3E4A89", # Blue
  "#31688E", # Blue-green
  "#26828E", # Turquoise
  "#1F9E89", # Green
  "#35B779", # Light green
  "#6DCD59", # Yellow-green
  "#B4DE2C", # Yellow
  "#FDE725", # Light yellow
  "#F9F9F9"  # Off-white for the last one
)

# Create a color factor without red
fisher_palette <- colorFactor(palette = viridis_colors, domain = unique(bcw_ground1_filtered$Fisher_ID))

# Start leaflet map
map <- leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%
 
 # Triangulated (non-flagged) points color-coded by Fisher_ID
addCircleMarkers(
  data = bcw_ground1_filtered %>% filter(Bearing_flagged != "flag"),
  lng = ~Longitude, lat = ~Latitude,
  color = ~fisher_palette(Fisher_ID),
  radius = 5,
  popup = ~paste("RowID:", row_id,
                 "<br>Fisher_ID:", Fisher_ID,
                 "<br>Triangulated:", Latitude, ",", Longitude),
  group = "Triangulated (By Fisher_ID)"
) %>%
  # # Orange triangulated points
  # addCircleMarkers(
  #   data = bcw_ground1_filtered %>% filter(Bearing_flagged != "flag"),
  #   lng = ~Longitude, lat = ~Latitude,
  #   color = "orange",
  #   radius = 5,
  #   popup = ~paste("RowID:", row_id, "<br>Fisher_ID: ", Fisher_ID, "<br>Triangulated:", Latitude, ",", Longitude),
  #   group = "Triangulated (Orange)"
  # ) %>%
  
  # Red flagged triangulated points
  addCircleMarkers(
    data = bcw_ground1_filtered %>% filter(Bearing_flagged == "flag"),
    lng = ~Longitude, lat = ~Latitude,
    color = "red",
    radius = 5,
    popup = ~paste("RowID:", row_id, "<br>Fisher_ID: ", Fisher_ID, "<br>FLAGGED:", Latitude, ",", Longitude),
    group = "Flagged (Red)"
  )%>%
  
  # Add legend for Fisher_ID
  addLegend(
    position = "bottomright",
    pal = fisher_palette,
    values = bcw_ground1_filtered$Fisher_ID,
    title = "Fisher_ID",
    opacity = 1
  )

# ---------- INSERT THIS UPDATED LOOP HERE ----------

# Loop through bearings 1–9
for (i in 1:9) {
  lat_col <- paste0("Bearing", i, "_lat")
  lon_col <- paste0("Bearing", i, "_long")
  bearing_col <- paste0("Bearing", i, "_bearing")
  
  # Bearings from flagged rows (gray)
  flagged_bearings <- bcw_ground1_filtered %>%
    filter(Bearing_flagged == "flag",
           !is.na(.data[[lat_col]]) & !is.na(.data[[lon_col]])) %>%
    mutate(
      lat = .data[[lat_col]],
      lon = .data[[lon_col]],
      bearing_val = .data[[bearing_col]]
    )
  
  map <- map %>%
    addCircleMarkers(
      data = flagged_bearings,
      lng = ~lon, lat = ~lat,
      color = "gray",
      radius = 5,
      popup = ~paste0("RowID: ", row_id,
                      "<br>Fisher_ID: ", Fisher_ID,
                      "<br>FLAGGED Bearing", i,
                      "<br>Location: ", round(lat, 5), ", ", round(lon, 5),
                      "<br>Bearing: ", bearing_val),
      group = "Flagged Bearings (Gray)"
    )
  
  # Bearings from non-flagged rows (black)
  regular_bearings <- bcw_ground1_filtered %>%
    filter(Bearing_flagged != "flag",
           !is.na(.data[[lat_col]]) & !is.na(.data[[lon_col]])) %>%
    mutate(
      lat = .data[[lat_col]],
      lon = .data[[lon_col]],
      bearing_val = .data[[bearing_col]]
    )
  
  map <- map %>%
    addCircleMarkers(
      data = regular_bearings,
      lng = ~lon, lat = ~lat,
      color = "black",
      radius = 5,
      popup = ~paste0("RowID: ", row_id,
                      "<br>Fisher_ID: ", Fisher_ID,
                      "<br>Bearing", i,
                      "<br>Location: ", round(lat, 5), ", ", round(lon, 5),
                      "<br>Bearing: ", bearing_val),
      group = "Bearings (Black)"
    )

}

# ---------------------------------------------------

# Add layer controls
map <- map %>%
  addLayersControl(
    overlayGroups = c(
      "Triangulated (By Fisher_ID)",
      "Flagged (Red)",
      "Bearings (Black)",
      "Flagged Bearings (Gray)"
    ),
    options = layersControlOptions(collapsed = FALSE)
  )

# Show map
map

# Save as interactive HTML
saveWidget(map, "bcw_ground1_map_with_rowid.html", selfcontained = TRUE)

# Drop all list-type columns
bcw_ground1_flat <- bcw_ground1 %>%
 dplyr::select(where(~ class(.) != "list"))

# Save to CSV
write.csv(bcw_ground1_flat, file = "BCWild_ground_bearingsextract.csv", row.names = FALSE)


```


12. Upload ground telemetry from Clappia - this binds to BC wild data and contains fine-scale habitat data 
standardize the columns & attach to BC wild ground

```{r}
clappia <- read.csv(file="Input_VHF_data_sheets/Clappia_Location_downloaded01May2025.csv", header=T)

clappia <- clappia %>%
  mutate(Date = as.Date(Date, format = "%b %d %Y"))

clappia$Join_ID <- paste0(clappia$Fisher_ID, " ", clappia$Date)

### bring in bcw_ground1 and make a join category
bcw_ground1$Join_ID <- paste0(bcw_ground1$Fisher_ID, " ", bcw_ground1$Date)

# Filter clappia and join to bcw_ground1
bcw_ground1 <- bcw_ground1 %>%
  left_join(
    clappia %>%dplyr::select(Join_ID, Surveyor, Tree_species, DBH),
    by = "Join_ID"
  )

names(bcw_ground1)

## Environmental covariates: "wind", "cloud_cover", "DBH" ,"Tree_species", "ambient_temperature", "Surveyor", "precip"

##Check fm_compile1 names
names(fm_compile1)

## Environmental covariates: "Temperature", "Surveyor", "Wind", "Precip", "Tree_species", "DBH", "Snow_depth", "Cloud_cover"

## rename the bcw_ground1 names to match

# Rename specific columns
bcw_ground1 <- bcw_ground1 %>%
  rename(
    Wind = wind,
    Temperature = ambient_temperature,
    Precip = precip, 
    Cloud_cover = cloud_cover
  )
```


13. Upload BC Wild Aerial telemetry and standardize column names

```{r}
bcw_aerial <- read.csv(file="Input_VHF_data_sheets/Aerial telemetry BC Wild-02may2025.csv")

# Rename specific columns
bcw_aerial<- bcw_aerial %>%
  rename(
    Latitude = latitude,
    Longitude = longitude,
    UTM_easting = easting, 
    UTM_northing = northing,
    Date= userDateTime,
    Frequency = frequency,
    Location_UUID = locationId
  )

bcw_aerial$Date <- as.Date(bcw_aerial$Date, "%Y-%m-%d")

bcw_aerial$Radiolocation_type <- "Aerial telemetry"

bcw_aerial$Data_source <- "BCWild Aerial Telemetry"

bcw_aerial$Precision_category <- "NA"
bcw_aerial$Precision_adjusted <- "4" #Home range level

## attach the Fisher IDs by frequency
freq <- read.csv(file="Fisher_frequency_list.csv")

freq_selected <- freq %>%
 dplyr::select(Frequency, Fisher_ID)

bcw_aerial_joined <- bcw_aerial %>%
  left_join(freq_selected, by = "Frequency")

bcw_aerial_joined$Error_area <- ""
bcw_aerial_joined$Error_northing <- ""
bcw_aerial_joined$Error_easting <- ""

# Convert the dataframe to an sf object
bcw_aerial_sf <- st_as_sf(bcw_aerial, coords = c("Longitude", "Latitude"), crs = 4326)

# Define a function to convert Latitude/Longitude to UTM (Zone 10)
lat_lon_to_utm <- function(lat, lon) {
  st_point(c(lon, lat)) %>%
    st_sfc(crs = 4326) %>%
    st_transform(crs = 32610) %>%
    st_coordinates() %>%
    as.vector()
}

# Apply the function to rows where UTM columns are NA
bcw_aerial <- bcw_aerial %>%
  rowwise() %>%
  mutate(
    UTM_easting = ifelse(is.na(UTM_easting), lat_lon_to_utm(Latitude, Longitude)[1], UTM_easting),
    UTM_northing = ifelse(is.na(UTM_northing), lat_lon_to_utm(Latitude, Longitude)[2], UTM_northing)
  )

```


14. Combine all datasets, extract the following columns:

Location_UUID
Fisher_ID
Date
UTM_easting
UTM_northing
Latitude
Longitude
Radiolocation_type
Precision_category
Precision_adjusted
Error_area
Error_easting
Error_northing
Data_source
Surveyor
Den_tree_name

```{r}
# Define the common columns

# rename filemaker utms
fm_compile1 <- fm_compile1 %>%
  rename(
    UTM_easting=UTM_Easting,
    UTM_northing = UTM_Northing
  )

#Add Snow depth to bcw_ground1
bcw_ground1$Snow_depth = ""

#Check names of the aerial files and add blank columns for environmental variables. 
names(bcw_aerial_joined)
bcw_aerial_joined <- bcw_aerial_joined %>%
  # rename(
  #   Surveyor = observer ) %>%
  mutate(Temperature = "",
  Precip = "",
  Cloud_cover = "",
  Wind = "",
  Tree_species= "",
  DBH = "",
  Snow_depth = ""
         )
names(aerial_caps)
aerial_caps <- aerial_caps %>%
  mutate(Temperature = "",
  Precip = "",
  Cloud_cover = "",
  Wind = "",
  Tree_species= "",
  DBH = "",
  Snow_depth = ""
         )

common_cols <- c(
  "Location_UUID",
  "Fisher_ID",
  "Date",
  "UTM_easting",
  "UTM_northing",
  "Latitude",
  "Longitude",
  "Radiolocation_type",
  "Precision_adjusted",
  "Precision_category",
  "Error_area",
  "Error_easting",
  "Error_northing",
  "Surveyor",
  "Data_source",
  "Temperature",
  "Precip",
  "Cloud_cover",
  "Wind",
  "Tree_species",
  "DBH",
  "Snow_depth"
 )

# List of data frames
dataframes <- list(fm_compile1, aerial_caps, bcw_ground1, bcw_aerial_joined)

# Convert Precision_adjusted to character and Error_area to numeric for all data frames
dataframes <- lapply(dataframes, function(df) {
  df %>%
    mutate(
      Precision_adjusted = as.character(Precision_adjusted),
      Precision_category = as.character(Precision_category),
      Error_area = as.numeric(Error_area),
      Error_easting = as.numeric(Error_easting),
      Error_northing = as.numeric(Error_northing),
      Temperature = as.character(Temperature),
      Cloud_cover = as.character(Cloud_cover),
      Date = as.Date(Date),
      Wind = as.character(Wind),
      DBH = as.character(DBH),
      Snow_depth = as.character(Snow_depth)
    )
})

# Assign the updated data frames back to their original names
fm_compile1 <- dataframes[[1]]
aerial_caps <- dataframes[[2]]
bcw_ground1 <- dataframes[[3]]
bcw_aerial_joined <- dataframes[[4]]

# # Filter to common columns
fm_compile1_filtered <- fm_compile1 %>% dplyr::select(all_of(common_cols))
aerial_caps_filtered <- aerial_caps %>% dplyr::select(all_of(common_cols))
bcw_ground_filtered <- bcw_ground1 %>% dplyr::select(all_of(common_cols))
bcw_aerial_filtered <- bcw_aerial_joined %>% dplyr::select(all_of(common_cols))


# Row bind all datasets
all_data <- bind_rows(
  fm_compile1_filtered,
  aerial_caps_filtered,
  bcw_ground_filtered,
  bcw_aerial_filtered
)



```

15. Add Shannon's denning data by column bind, and by row bind for new data
```{r}
dens <- read.csv("Input_VHF_data_sheets/Enterprise Fisher Den Trees.csv", header=T)

dens$Date <- str_extract(dens$Den_Tree_Name, "\\d{4}-\\d{2}-\\d{2}")
dens$Date <- as.Date(dens$Date)

all_data$Radiolocation_type <- ifelse(all_data$Radiolocation_type=="Aerial", "Aerial telemetry", all_data$Radiolocation_type)

# Perform the left join and update Tree_species
# Filter all_data to exclude rows with Radiolocation_type == "Aerial telemetry"
all_data_filtered <- all_data %>%
  filter(Radiolocation_type != "Aerial telemetry")

# Perform the left join with dens
Merged_data <- all_data_filtered%>%
  left_join(
    dens %>% dplyr::select(Fisher_ID, Date, Den_Tree_Name, Tree_species),
    by = c("Fisher_ID", "Date")
  ) %>%
  mutate(
    Tree_species = coalesce(Tree_species.x, Tree_species.y)
  ) %>%
  dplyr::select(-Tree_species.x, -Tree_species.y)

# If you want to include the rows with Radiolocation_type == "Aerial telemetry" without joining dens
# and keep their original Tree_species, you can bind them back
aerial_data <- all_data %>%
  filter(Radiolocation_type == "Aerial telemetry")

# Combine the datasets
all_data1 <- bind_rows(Merged_data, aerial_data)


dens1 <- dens %>% filter(Bind =="n")%>%
  mutate(
  Surveyor = "Shannon Werden", 
  Location_UUID ="",
  Radiolocation_type = "Den tree",
  Data_source = "Enterprise fisher Den trees",
  Precision_category = 0,
  Precision_adjusted = 0,
  Error_area = "",
  Error_easting = "",
  Error_northing = "",
  Temperature = "",
  Precip = "",
  Cloud_cover = "",
  Wind = "",
  DBH = "",
  Snow_depth = ""
         )

# Step 1: Convert dens1 to an sf object using Latitude and Longitude
dens1_sf <- st_as_sf(dens1, coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)

# Step 2: Transform coordinates to UTM Zone 10N (EPSG:32610)
dens1_utm <- st_transform(dens1_sf, crs = 32610)

# Step 3: Extract UTM easting and northing
utm_coords <- st_coordinates(dens1_utm)
dens1_utm$UTM_easting <- utm_coords[, 1]
dens1_utm$UTM_northing <- utm_coords[, 2]

# Step 4: Remove geometry column if not needed
dens1 <- st_drop_geometry(dens1_utm)

dens1 <- dens1 %>% dplyr::select(-Bind)

Merged_data1 <- rbind(all_data1, dens1)

# Make sure all confirmed den trees are marked as Radiolocation_Type "Den tree"
Merged_data1 <- Merged_data1 %>%
  mutate(Radiolocation_type = if_else(!is.na(Den_Tree_Name), "Den tree", Radiolocation_type))

vhf_cleaned <- Merged_data1
```


16. Fix UTMs for all data, Remove additional duplicates by individual, datetime, and location

```{r}

 # vhf_cleaned <- read.csv(file="Fisher_VHF_locations_May2025.csv", header=T)


### There were missing UTMs in the datasheet so  I have this code to populate from Lat/Long
# 
# ## Fix missing UTMs
# missing_utm <- vhf_cleaned[is.na(vhf_cleaned$UTM_easting) | is.na(vhf_cleaned$UTM_northing), ]
# # Create an sf object with the missing rows
# missing_sf <- st_as_sf(missing_utm, coords = c("Longitude", "Latitude"), crs = 4326)
# 
# # Transform to UTM Zone 10N (EPSG: 32610)
# missing_sf_utm <- st_transform(missing_sf, crs = 32610)
# 
# # Extract UTM coordinates
# missing_utm_coords <- st_coordinates(missing_sf_utm)
# 
# # Add the UTM coordinates back to the original dataframe
# vhf_cleaned$UTM_easting[is.na(vhf_cleaned$UTM_easting)] <- missing_utm_coords[, 1]
# vhf_cleaned$UTM_northing[is.na(vhf_cleaned$UTM_northing)] <- missing_utm_coords[, 2]

## I have since discovered that UTMs assembled in filemaker contain errors and are not aligned with the lat/long values, which appear to be correct. This is important for KDE estimation later, requiring UTMs. 
## Therefore, I am overwriting UTMs with missing and erroenous values with a conversion from Lat/Long

vhf_cleaned_sf <- st_as_sf(vhf_cleaned, coords = c("Longitude", "Latitude"), crs = 4326)

vhf_cleaned_utm <- st_transform(vhf_cleaned_sf, crs = 32610)

utm_coords <- st_coordinates(vhf_cleaned_utm)

vhf_cleaned$UTM_easting <- utm_coords[, 1]
vhf_cleaned$UTM_northing <- utm_coords[, 2]


## Previous code to remove duplicates based on the Radiolocation_type

# # Remove duplicates with preference for 'Element' in Radiolocation_type
# vhf_cleaned <- vhf_cleaned %>%
#   # Arrange to prioritize 'Element' entries
#   arrange(Fisher_ID, Date,
#           desc(Radiolocation_type == "Element")) %>%
#   # Remove duplicates based on specified columns
#   distinct(Fisher_ID, UTM_northing, UTM_easting, .keep_all = TRUE) ## dropped 2 locations

## or by date and not location in case locations are same spot but off in different rows
vhf_cleaned <- vhf_cleaned %>%
  # Arrange to prioritize 'Element' entries
  arrange(Fisher_ID, Date,
          desc(Radiolocation_type == "Element")) %>%
  # Remove duplicates based on specified columns
  distinct(Fisher_ID, Date, .keep_all = TRUE) ## dropped 41 locations
```


17. Precision category cleaning, removing outliers and deletemes

```{r}

vhf_cleaned <- vhf_cleaned %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    row_id = row_number()
  )
## Remove Precision_adjusted= "discard"
vhf_cleaned <- vhf_cleaned %>%
  filter(Precision_adjusted != "discard",
         Fisher_ID !="Deleteme", Fisher_ID!="Deleteme2", Fisher_ID!="", Fisher_ID !="CF12") ## removed 5 more locations

vhf_cleaned <- vhf_cleaned %>%
  group_by(Fisher_ID) %>%
  mutate(
    centroid_lat = mean(Latitude, na.rm = TRUE),
    centroid_lon = mean(Longitude, na.rm = TRUE),
    distance_to_centroid = distGeo(cbind(Longitude, Latitude), cbind(centroid_lon, centroid_lat)),
    z_score = scale(distance_to_centroid)[,1],  # Convert to numeric
    Flagged = ifelse(abs(z_score) > 3, TRUE, FALSE)  # Flag points >3 SDs away
  ) %>%
  ungroup() 

## Remove flagged locations
vhf_cleaned <- subset(vhf_cleaned, vhf_cleaned$Flagged !="TRUE" | is.na(vhf_cleaned$Flagged)=="TRUE") #dropped 12 locations

```

18. Check sample sizes by Fisher_ID and Radiolocation_type, export

```{r}
table(vhf_cleaned$Fisher_ID) #517 records
#F02, FO3, F04, F07, F07, F09 all have 50-72 locations - remove the deleteme records


vhf_cleaned <- subset(vhf_cleaned, vhf_cleaned$Fisher_ID !="Deleteme" & vhf_cleaned$Fisher_ID !="Deleteme2" )

## combine Aerial and Aerial telemetry into one category

vhf_cleaned$Radiolocation_type <- ifelse(vhf_cleaned$Radiolocation_type=="Aerial", "Aerial telemetry", vhf_cleaned$Radiolocation_type)

## Fisher_ID by location type

freq_table <- table(vhf_cleaned$Fisher_ID, vhf_cleaned$Radiolocation_type)

write.csv(freq_table, file="Fisher_ID_by_Radiolocation_type.csv")
```


19. Map the data to look for any outliers visually:
```{r}
# Ensure Latitude and Longitude are numeric and row_id exists
vhf_cleaned <- vhf_cleaned %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    row_id = row_number()
  )

# Create a viridis color palette for Fisher_IDs
viridis_colors <- c(
  "#440154", "#482878", "#3E4A89", "#31688E", "#26828E", 
  "#1F9E89", "#35B779", "#6DCD59", "#B4DE2C", "#FDE725", "#F9F9F9"
)

fisher_palette <- colorFactor(
  palette = viridis_colors,
  domain = unique(vhf_cleaned$Fisher_ID)
)

# Add a column for circle color
vhf_cleaned <- vhf_cleaned %>%
  mutate(
    circle_color = ifelse(Flagged == TRUE, "red", fisher_palette(Fisher_ID))
  )

# Create the leaflet map
vhf_map <- leaflet(vhf_cleaned) %>%
  addProviderTiles("OpenStreetMap") %>%
  addCircleMarkers(
    lng = ~Longitude, lat = ~Latitude,
    color = ~circle_color,
    radius = 5,
    popup = ~paste0(
      "RowID: ", row_id,
      "<br>Fisher_ID: ", Fisher_ID,
      "<br>Radiolocation_type: ", Radiolocation_type,
      "<br>Precision_adjusted: ", Precision_adjusted
    ),
    group = "VHF Locations"
  ) %>%
  addLegend(
    position = "bottomright",
    pal = fisher_palette,
    values = ~Fisher_ID,
    title = "Fisher_ID",
    opacity = 1
  ) %>%
  addLayersControl(
    overlayGroups = c("VHF Locations"),
    options = layersControlOptions(collapsed = FALSE)
  )

# Display the map
vhf_map

# Optionally save it
saveWidget(vhf_map, "vhf_cleaned_map.html", selfcontained = TRUE)


# Optionally save it
saveWidget(vhf_map, "vhf_cleaned_map.html", selfcontained = TRUE)

```



20. Run a prelim AKDE on the fisher locations and map with the den trees in red.

```{r}

# Load and preprocess data
vhf_cleaned <- read.csv(file = "Fisher_VHF_locations_May2025.csv", header = TRUE)

vhf_cleaned <- vhf_cleaned %>%
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    Date = as.Date(Date)  # Ensure Date is in Date format
  )

# Identify Fisher_IDs with sufficient data for home range calculation
vhf_kde_data <- vhf_cleaned %>%
  filter(!is.na(UTM_easting), !is.na(UTM_northing)) %>%
  group_by(Fisher_ID) %>%
  filter(n() >= 15) %>%
  ungroup()

# Identify Fisher_IDs without sufficient data
vhf_non_kde_data <- vhf_cleaned %>%
  filter(!(Fisher_ID %in% unique(vhf_kde_data$Fisher_ID)))

# Convert to sf objects
vhf_sf <- st_as_sf(vhf_kde_data, coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)
vhf_sf_utm <- st_transform(vhf_sf, crs = 32610)

# Convert to Spatial object for kernelUD
vhf_sp <- as(vhf_sf_utm, "Spatial")
row.names(vhf_sp) <- paste0(vhf_sp$Fisher_ID, "_", seq_len(nrow(vhf_sp)))

# Calculate kernel utilization distribution
kud <- kernelUD(vhf_sp[, "Fisher_ID"], h = "href")

# Extract 95% home range polygons
hr_95 <- getverticeshr(kud, percent = 95)
hr_sf <- st_as_sf(hr_95) %>%
  st_transform(crs = 4326)

# Create color palette
fisher_ids <- unique(vhf_cleaned$Fisher_ID)
fisher_palette <- colorFactor(palette = viridis(length(fisher_ids)), domain = fisher_ids)

# Prepare main VHF points
vhf_sf <- vhf_sf %>%
  mutate(circle_color = ifelse(Flagged == TRUE, "red", fisher_palette(Fisher_ID)))

# Prepare non-KDE VHF points
vhf_non_kde_sf <- st_as_sf(vhf_non_kde_data, coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE) %>%
  mutate(circle_color = ifelse(Flagged == TRUE, "red", fisher_palette(Fisher_ID)))

# Prepare Den tree points
den_tree_sf <- vhf_cleaned %>%
  filter(Radiolocation_type == "Den tree") %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326, remove = FALSE)

# Assign colors to home range polygons
hr_sf <- hr_sf %>%
  mutate(color = fisher_palette(id))

# Initialize the leaflet map
leaflet_map <- leaflet() %>%
  addProviderTiles("OpenStreetMap") %>%
  
  # Add VHF points with home ranges
  addCircleMarkers(
    data = vhf_sf,
    lng = ~Longitude,
    lat = ~Latitude,
    color = ~circle_color,
    radius = 5,
    popup = ~paste0(
      "<b>Fisher_ID:</b> ", Fisher_ID,
      "<br><b>Date:</b> ", Date,
      "<br><b>Radiolocation_type:</b> ", Radiolocation_type,
      "<br><b>Precision_adjusted:</b> ", Precision_adjusted,
      "<br><b>Latitude:</b> ", Latitude,
      "<br><b>Longitude:</b> ", Longitude
    ),
    group = "VHF Locations"
  ) %>%
  
  # Add VHF points without home ranges
  addCircleMarkers(
    data = vhf_non_kde_sf,
    lng = ~Longitude,
    lat = ~Latitude,
    color = ~circle_color,
    radius = 5,
    popup = ~paste0(
      "<b>Fisher_ID:</b> ", Fisher_ID,
      "<br><b>Date:</b> ", Date,
      "<br><b>Radiolocation_type:</b> ", Radiolocation_type,
      "<br><b>Precision_adjusted:</b> ", Precision_adjusted,
      "<br><b>Latitude:</b> ", Latitude,
      "<br><b>Longitude:</b> ", Longitude
    ),
    group = "VHF Locations (No Home Range)"
  ) %>%
  
  # Add Den tree locations in red
  addCircleMarkers(
    data = den_tree_sf,
    lng = ~Longitude,
    lat = ~Latitude,
    color = "red",
    radius = 6,
    popup = ~paste0(
      "<b>Fisher_ID:</b> ", Fisher_ID,
      "<br><b>Date:</b> ", Date,
      "<br><b>Den_Tree_Name:</b> ", Den_Tree_Name,
      "<br><b>Precision_adjusted:</b> ", Precision_adjusted,
      "<br><b>Latitude:</b> ", Latitude,
      "<br><b>Longitude:</b> ", Longitude
    ),
    group = "Den Trees"
  ) %>%
  
  # Add home range polygons
  addPolygons(
    data = hr_sf,
    color = ~color,
    weight = 2,
    fillOpacity = 0.3,
    popup = ~paste0("<b>Fisher_ID:</b> ", id),
    group = "Home Ranges"
  ) %>%
  
  # Add legend
  addLegend(
    position = "bottomright",
    pal = fisher_palette,
    values = fisher_ids,
    title = "Fisher_ID",
    opacity = 1
  ) %>%
  
  # Add layer controls
  addLayersControl(
    overlayGroups = c("VHF Locations", "VHF Locations (No Home Range)", "Den Trees", "Home Ranges"),
    options = layersControlOptions(collapsed = FALSE)
  )

# Display the map
leaflet_map

#Export map to a new folder
if (!dir.exists("html_maps")) {
  dir.create("html_maps")
}
saveWidget(leaflet_map, file = "html_maps/VHF_cleaned_HR_map.html", selfcontained = TRUE)
```


21. Export final data and home range shapefiles

```{r}

#Export the data
write.csv(vhf_cleaned, file="Fisher_VHF_locations_May2025.csv")

##Save the shapefiles
if (!dir.exists("95_KDE_Shapefiles")) {
  dir.create("95_KDE_Shapefiles")
}
st_write(hr_sf, dsn = "95_KDE_Shapefiles/home_ranges_95.shp", delete_layer = TRUE)

```

